---
title: "Reproducible Research Week 4 Project"
author: "Siddhesh Panhale"
date: "8/12/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Introduction**

-Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

-This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

-The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events.

**Synopsis**

-This is an exploration of the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database.

-This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, which type of event, as well as the estimates of relevant fatalities, injuries, and various forms of damage.

-The dataset used in this project is provided by the U.S. National Oceanic and Atmospheric Administration (NOAA).

-This analysis discovered that tornados are responsible for a maximum number of fatalities and injuries.

-This analysis also discoered that floods are responsbile for maximum property damage, while Droughts cause maximum crop damage.

**Data Processing**

Loading Libraries
```{r , results="hide"}
library(knitr)
library(plyr)
library(ggplot2)
library(lattice)
library(data.table)
library(grid)
library(gridExtra)
library(R.utils)
library(rmarkdown)
```

Downloading, Uncompressing and Loading the file
```{r}
temp <- tempfile()

if(!file.exists("/stormData.csv.bz2")){
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile="./stormData.csv.bz2")
}
if(!file.exists("stormdata.csv"))
{
  bunzip2("stormData.csv.bz2","stormdata.csv",remove=F)
}

storm <- read.csv("stormdata.csv",header=TRUE,sep=",")
summary(storm)
names(storm)

variables<-c("EVTYPE","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP","CROPDMG","CROPDMGEXP")
strmdata<-storm[variables]

dim(strmdata)
names(strmdata)
```

**Results**

**1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?**

Reviewing events that cause the most fatalities
```{r}
Fatalities <- aggregate(FATALITIES ~ EVTYPE, data = strmdata, FUN = sum)
Top10_Fatalities <- Fatalities[order(-Fatalities$FATALITIES), ][1:10, ] 
Top10_Fatalities 
```

Reviewing events that cause the most injuries
```{r}
Injuries <- aggregate(INJURIES ~ EVTYPE, data = strmdata, FUN = sum)
Top10_Injuries <- Injuries[order(-Injuries$INJURIES), ][1:10, ] 
Top10_Injuries
```

Plot of Top 10 Fatalities & Injuries for Weather Event Types
```{r}
par(mfrow=c(1,2),mar=c(10,3,3,2))
barplot(Top10_Fatalities$FATALITIES,names.arg=Top10_Fatalities$EVTYPE,las=2,col="red",ylab="fatalities",main="Top 10 fatalities")
barplot(Top10_Injuries$INJURIES,names.arg=Top10_Injuries$EVTYPE,las=2,col="red",ylab="injuries",main="Top 10 Injuries")
```

**2. Across the United States, which types of events have the greatest economic consequences?**

Getting Property damage
```{r}
unique(strmdata$PROPDMGEXP)
```

Assigning values for the property exponent and setting 0 to invalid exponent strmdata
```{r}
strmdata$PROPEXP[strmdata$PROPDMGEXP == "K"] <- 1000
strmdata$PROPEXP[strmdata$PROPDMGEXP == "M"] <- 1e+06
strmdata$PROPEXP[strmdata$PROPDMGEXP == ""] <- 1
strmdata$PROPEXP[strmdata$PROPDMGEXP == "B"] <- 1e+09
strmdata$PROPEXP[strmdata$PROPDMGEXP == "m"] <- 1e+06
strmdata$PROPEXP[strmdata$PROPDMGEXP == "0"] <- 1
strmdata$PROPEXP[strmdata$PROPDMGEXP == "5"] <- 1e+05
strmdata$PROPEXP[strmdata$PROPDMGEXP == "6"] <- 1e+06
strmdata$PROPEXP[strmdata$PROPDMGEXP == "4"] <- 10000
strmdata$PROPEXP[strmdata$PROPDMGEXP == "2"] <- 100
strmdata$PROPEXP[strmdata$PROPDMGEXP == "3"] <- 1000
strmdata$PROPEXP[strmdata$PROPDMGEXP == "h"] <- 100
strmdata$PROPEXP[strmdata$PROPDMGEXP == "7"] <- 1e+07
strmdata$PROPEXP[strmdata$PROPDMGEXP == "H"] <- 100
strmdata$PROPEXP[strmdata$PROPDMGEXP == "1"] <- 10
strmdata$PROPEXP[strmdata$PROPDMGEXP == "8"] <- 1e+08

strmdata$PROPEXP[strmdata$PROPDMGEXP == "+"] <- 0
strmdata$PROPEXP[strmdata$PROPDMGEXP == "-"] <- 0
strmdata$PROPEXP[strmdata$PROPDMGEXP == "?"] <- 0

strmdata$PROPDMGVAL <- strmdata$PROPDMG * strmdata$PROPEXP
```

Getting Crop damage
```{r}
unique(strmdata$CROPDMGEXP)
```

Assigning values for the crop exponent and setting 0 to invalid exponent strmdata
```{r}
strmdata$CROPEXP[strmdata$CROPDMGEXP == "M"] <- 1e+06
strmdata$CROPEXP[strmdata$CROPDMGEXP == "K"] <- 1000
strmdata$CROPEXP[strmdata$CROPDMGEXP == "m"] <- 1e+06
strmdata$CROPEXP[strmdata$CROPDMGEXP == "B"] <- 1e+09
strmdata$CROPEXP[strmdata$CROPDMGEXP == "0"] <- 1
strmdata$CROPEXP[strmdata$CROPDMGEXP == "k"] <- 1000
strmdata$CROPEXP[strmdata$CROPDMGEXP == "2"] <- 100
strmdata$CROPEXP[strmdata$CROPDMGEXP == ""] <- 1

strmdata$CROPEXP[strmdata$CROPDMGEXP == "?"] <- 0

strmdata$CROPDMGVAL <- strmdata$CROPDMG * strmdata$CROPEXP
```

Property Damage Summary

Procedure = aggregate the property damage by the event type and sort the output it in descending order
```{r}
prop <- aggregate(PROPDMGVAL~EVTYPE,data=strmdata,FUN=sum,na.rm=TRUE)
prop <- prop[with(prop,order(-PROPDMGVAL)),]
prop <- head(prop,10)
print(prop)
```

Crop Damage Summary
```{r}
crop <- aggregate(CROPDMGVAL~EVTYPE,data=strmdata,FUN=sum,na.rm=TRUE)
crop <- crop[with(crop,order(-CROPDMGVAL)),]
crop <- head(crop,10)
print(crop)
```

Graph showing the top 10 property and crop damages
```{r}
par(mfrow=c(1,2),mar=c(11,3,3,2))
barplot(prop$PROPDMGVAL/(10^9),names.arg=prop$EVTYPE,las=2,col="Red",ylab="Prop.damage(billions)",main="Top10 Prop.Damages")
barplot(crop$CROPDMGVAL/(10^9),names.arg=crop$EVTYPE,las=2,col="Red",ylab="Crop damage(billions)",main="Top10 Crop.Damages")
```

**Conclusion**

As seen from the graph, we can say that Tornado is responsible for the maximum number of fatalities and injuries while Floods are responsbile for maximum property damage and Droughts cause maximum crop damage.
